# Lab 3
---
title: "Lab 3"
author: "Lucas Finke"
date: "2023-09-18"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, echo= FALSE}
setwd("C:/Users/Lucas/Downloads/Kuhnert+Venables-R_Course_Notes/Session1/Session1/")
source("FrontMatter.R")
```
Intalling packages.

```{R Animals ALL}
setwd("C:/Users/Lucas/Downloads/Kuhnert+Venables-R_Course_Notes/Session1/Session1/")
read.csv("df.RData")
attach(AnimalsALL)
plot(body,brain,xlab="Average body weight (kg)", ylab="Average brain weight (gm)")
plot(body,brain,xlab="Average body weight (kg)", ylab= "Average brain weight (gm)", log="xy")
identify(body,brain,row.names(AnimalsALL))
plot(log(body),log(brain),pch=4)
abline(lsfit(log(body),log(brain)),lty=1)
abline(ltsreg(log(body),log(brain)),lty=2)
legend(locator(1),legend=c("OLS line","LTS line"),lty=1:2)
```
Using AnimalsALL from the previous lab, comparing the results in a log scale in both directions.Then identifying which Animals happen to be outliers in the regression, which happen to be dinosaurs. After plotting, a legend was created using a locator.


```{R Survey Data}
require(MASS)
attach(survey)
plot(survey)
plot(Exer)
table(Exer)
names(Exer) <- c("Freq","None","Some")
pie(as.integer(c(115,24,98)),labels=c("Freq","None","Some"),col=1:3)
legend("topright",c("Freq","None","Some"),fill=1:3)
```

Using Data from Adelaide University's first year students survey results to create a table, plot, and pie chart with their responses to see which one provides a better visual of the data. From there, a legend with the responses were made. 

```{R Banknote}
require('rtools43')
install.packages("remotes")
installed.packages('rtools43')
remotes::install_github("cran/alr3")
```
```{r}
require(alr3)
data(banknote)
attach(banknote)
pairs(banknote)
pairs(banknote[,-1],panel=
        function(x,y,fake) {
xy <- cbind(x,y)
points(xy[fake==0,],pch=15)
points(xy[fake==1,],pch=0)
}, fake=Y)
coplot(var1 ~ var2 | Y,data=banknote)
```

Using the old package of alr3 to detect counterfeit banknotes and plotting the lengths to separate the fakes. Then using a coplot to see if the legal notes are undected.
