---
title: "Multiway Data"
author: "Song Qian"
date: "11/14/2022"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
source("FrontMatter.R")
packages(lattice)
packages(foreign)
packages(gdata)
data.restore (paste(dataDIR, "visualizing.data", sep="/"))

wt.bisquare <- function(u, c=6) {
 ifelse( abs(u/c) < 1, (1-(u/c)^2)^2, 0)
} ## bi-square weights

```

## Multiway data
The livestock count data (1987): collected for estimating European ammonia emission because animal source of ammonia accounts for 81% of all European emission. 

The data
```{r}
    attach(livestock)
    logcount <- log10(count)
    new.country <- ordered(country, rev(sort(levels(country))))
    new.livestock <- ordered(livestock.type,                c("Sheep","Poultry","Pigs","Horses","Cattle"))
    ans <- dotplot(new.country ~ logcount | new.livestock,
                   xlab = "Log 10 Number of Livestock",
                   layout=c(3,2))
    detach()
    ans
```

Another look -- changing the visual perception
```{R}
    attach(livestock)
    logcount <- log10(count)
    new.country <- reorder.factor(country, logcount, median)
    new.livestock <- reorder.factor(livestock.type, logcount, median)
    ans <- dotplot(new.country ~ logcount | new.livestock,
                   xlab = "Log 10 Number of Livestock",
                   layout=c(3,2))
    detach()
    ans
```

Two categorical variables and the visual perception of effects of them are asymmetric -- we can better compare values within a panel than values between panels.    

A different comparison
```{R}
    attach(livestock)
    logcount <- log10(count)
    new.country <- reorder.factor(country, logcount, median)
    new.livestock <- reorder.factor(livestock.type, logcount, median)
    ans <- dotplot(new.livestock ~ logcount | new.country,
                   xlab = "Log 10 Number of Livestock",
                   layout=c(4,7),
                   aspect=1/2)
    detach()
    ans
```


```{R}
    attach(livestock)
    logcount <- log10(count)
    wt <- rep(1,length(logcount))
    for(i in 1:10){
        liv.lm <- lm(logcount~livestock.type+country,weights=wt)
        wt <- wt.bisquare(liv.lm$res/median(abs(liv.lm$res)),c=6)
    }
    liv.effects <- dummy.coef(liv.lm)
    liv.effects <- sapply(liv.effects,sort)
    all.effects <- c(liv.effects$livestock.type,NA,liv.effects$country)
    ans <- dotplot(all.effects,
                   aspect=1,
                   xlab = "Log Livestock Effects (log 10 count)")
    detach()
    ans
    rm(wt)
```

### Additive fit
What we have examined pointed us to a more quantitative model:
$$ y_{lc} = \mu+\alpha_l+\beta_c + \epsilon_{lc}$$.
We want to know, the contribution of country and of livestock type. We redraw the previous two figures by ordering the countries and livestock type by their estimated effects.
```{R}
    attach(livestock)
    logcount <- log10(count)
    wt <- rep(1,length(logcount))
    for(i in 1:10){
        liv.lm <- lm(logcount~livestock.type+country,weights=wt)
        wt <- wt.bisquare(liv.lm$res/median(abs(liv.lm$res)),c=6)
    }
    liv.effects <- dummy.coef(liv.lm)
    liv.effects <- sapply(liv.effects,"sort")
    new.country <- ordered(country,names(liv.effects$country))
    new.livestock <- ordered(livestock.type,names(liv.effects$livestock.type))
    ans <- dotplot(new.country ~ logcount | new.livestock,
                   xlab = "Log 10 Number of Livestock",
                   layout=c(3,2))
    detach()
    ans
```
and
```{R}

    attach(livestock)
    logcount <- log10(count)
    wt <- rep(1,length(logcount))
    for(i in 1:10){
        liv.lm <- lm(logcount~livestock.type+country,weights=wt)
        wt <- wt.bisquare(liv.lm$res/median(abs(liv.lm$res)),c=6)
    }
    liv.effects <- dummy.coef(liv.lm)
    liv.effects <- sapply(liv.effects,"sort")
    new.country <- ordered(country,names(liv.effects$country))
    new.livestock <- ordered(livestock.type,names(liv.effects$livestock.type))
    ans <- dotplot(new.livestock ~ logcount | new.country,
                   xlab = "Log 10 Number of Livestock",
                   layout=c(4,7),
                   aspect=1)
    detach()
    ans
```
Fitted values of the log counts:     
```{R}
    attach(livestock)
    logcount <- log10(count)
    wt <- rep(1,length(logcount))
    for(i in 1:10){
        liv.lm <- lm(logcount~livestock.type+country,weights=wt)
        wt <- wt.bisquare(liv.lm$res/median(abs(liv.lm$res)),c=6)
    }
    liv.effects <- dummy.coef(liv.lm)
    liv.effects <- sapply(liv.effects,"sort")
    new.country <- ordered(country,names(liv.effects$country))
    new.livestock <- ordered(livestock.type,names(liv.effects$livestock.type))
    ans <- dotplot(new.country ~ fitted.values(liv.lm) | new.livestock,
                   xlab = "Fitted Log 10 Number of Livestock",
                   layout=c(3,2))
    detach()
    ans
```
and 
```{R}
    attach(livestock)
    logcount <- log10(count)
    wt <- rep(1,length(logcount))
    for(i in 1:10){
        liv.lm <- lm(logcount~livestock.type+country,weights=wt)
        wt <- wt.bisquare(liv.lm$res/median(abs(liv.lm$res)),c=6)
    }
    liv.effects <- dummy.coef(liv.lm)
    liv.effects <- sapply(liv.effects,"sort")
    new.country <- ordered(country,names(liv.effects$country))
    new.livestock <- ordered(livestock.type,names(liv.effects$livestock.type))
    ans <- dotplot(new.livestock ~ fitted.values(liv.lm) | new.country,
                   xlab = "Fitted Log 10 Number of Livestock",
                   layout=c(4,7),
                   aspect=1)
    detach()
    ans
```

Residuals:
```{R}
	attach(livestock)
	logcount <- log(count,2)
	wt <- rep(1,length(logcount))
	for(i in 1:10){
	        liv.lm <- lm(logcount~livestock.type+country,weights=wt)
	        wt <- wt.bisquare(liv.lm$res/median(abs(liv.lm$res)),c=6)
	}
	liv.effects <- dummy.coef(liv.lm)
	liv.effects <- sapply(liv.effects,"sort")
	new.country <- ordered(country,names(liv.effects$country))
	new.livestock <- ordered(livestock.type,names(liv.effects$livestock.type))
	liv.res <- residuals(liv.lm)
	liv.res[(livestock.type=="Pigs")&(country=="Turkey")] <- NA
	ans <- dotplot(new.country ~ liv.res | new.livestock,
		panel = function(x, y) {
			panel.dotplot(x, y)
	                panel.abline(v=0)
		},
		layout=c(3,2),
		xlab = "Residual Log 2 Number of Livestock")
	detach()
	ans
```
and
```{R}
    attach(livestock)
    logcount <- log(count,2)
    wt <- rep(1,length(logcount))
    for(i in 1:10){
        liv.lm <- lm(logcount~livestock.type+country,weights=wt)
        wt <- wt.bisquare(liv.lm$res/median(abs(liv.lm$res)),c=6)
    }
    liv.effects <- dummy.coef(liv.lm)
    liv.effects <- sapply(liv.effects,"sort")
    new.country <- ordered(country,names(liv.effects$country))
    new.livestock <- ordered(livestock.type,names(liv.effects$livestock.type))
    liv.res <- residuals(liv.lm)
    liv.res[(livestock.type=="Pigs")&(country=="Turkey")] <- NA
    ans <- dotplot(new.livestock ~ liv.res | new.country,
                   panel = function(x, y) {
                       panel.dotplot(x, y)
                       panel.abline(v=0)
                   },
                   aspect=1,
                   layout=c(4,7),
                   sub = list("Figure 6.10",cex=.8),
                   xlab = "Residual Log 2 Number of Livestock")
    detach()
    ans
```
## The Barley data:
```{R}
	dotplot(variety ~ yield | year * site,
		data = barley,
		aspect=.4,
		xlab = "Barley Yield (bushels/acre)")
```
We know that Morris was problematic. If we switch the year labeling for Morris:
```{R}
    attach(barley)
    morris31 <- yield[(site=="Morris")&(year=="1931")]
    morris32 <- yield[(site=="Morris")&(year=="1932")]
    new.yield <- yield
    new.yield[(site=="Morris")&(year=="1931")] <- morris32
    new.yield[(site=="Morris")&(year=="1932")] <- morris31
    ans <- dotplot(variety ~ new.yield | year * site, 
                   xlab = "Barley Yield (bushels/acre)",
                   aspect=.4)
    detach()
    ans
```
Displaying yearly differences: 1931 yield minus 1932
```{R}
    attach(barley)
    morris31 <- yield[(site=="Morris")&(year=="1931")]
    morris32 <- yield[(site=="Morris")&(year=="1932")]
    new.yield <- yield
    new.yield[(site=="Morris")&(year=="1931")] <- morris32
    new.yield[(site=="Morris")&(year=="1932")] <- morris31
    which <- year=="1931"
    new.yield.diff <- new.yield[which]-new.yield[!which]
    ans <- dotplot(variety[which] ~ new.yield.diff |
                       site[which], 
                   xlab = "Differences of Barley Yield (bushels/acre)",
                   layout = c(1,6),
                   aspect=.4)
    detach()
    ans
```
There is an interaction between year and site: not all sites had smaller yields in 1932
```{R}
    attach(barley)
    morris31 <- yield[(site=="Morris")&(year=="1931")]
    morris32 <- yield[(site=="Morris")&(year=="1932")]
    new.yield <- yield
    new.yield[(site=="Morris")&(year=="1931")] <- morris32
    new.yield[(site=="Morris")&(year=="1932")] <- morris31
    wt <- rep(1,length(yield))
    for(i in 1:10){
        barley.lm <- lm(new.yield~variety+year*site,weights=wt)
        wt <- wt.bisquare(barley.lm$res/median(abs(barley.lm$res)),c=6)
    }
    ans <- rfs(barley.lm, 
               aspect=2,
               ylab = "Yield (bushels/acre)")
    detach()
    ans
```
The year-site effects:
```{R}
    attach(barley)
    morris31 <- yield[(site=="Morris")&(year=="1931")]
    morris32 <- yield[(site=="Morris")&(year=="1932")]
    new.yield <- yield
    new.yield[(site=="Morris")&(year=="1931")] <- morris32
    new.yield[(site=="Morris")&(year=="1932")] <- morris31
    wt <- rep(1,length(yield))
    for(i in 1:10){
        barley.lm <- lm(new.yield~variety+year*site,weights=wt)
        wt <- wt.bisquare(barley.lm$res/median(abs(barley.lm$res)),c=6)
    }
    barley.effects <- dummy.coef(barley.lm)
    ys.effects <- c(barley.effects$"year:site" +
                    outer(barley.effects$year,barley.effects$site,"+"))
    ys.year <- ordered(rep(levels(year),6),levels(year))
    ys.site <- ordered(rep(levels(site),rep(2,6)),levels(site))
    n <- length(levels(ys.year))
    ans <- dotplot(ys.site ~ ys.effects,
                   ## panel = function(x, y, subscripts, ...) {
                   ##     do.call("abline", c(list(h = y), trellis.par.get("dot.line")))
                   ##	panel.superpose(x, y, subscripts, ...)
                                        #},
                   groups = ys.year,
                   aspect = 2/3,
                   xlab = "Site by Year Effects (bushels/acre)",
                   key = list(
                       points = Rows(trellis.par.get("superpose.symbol"), 1:n),
                       text = list(levels(ys.year)),
                       columns = n))
    detach()
    ans
```
The veriety effects:
```{R}
    attach(barley)
    morris31 <- yield[(site=="Morris")&(year=="1931")]
    morris32 <- yield[(site=="Morris")&(year=="1932")]
    new.yield <- yield
    new.yield[(site=="Morris")&(year=="1931")] <- morris32
    new.yield[(site=="Morris")&(year=="1932")] <- morris31
    wt <- rep(1,length(yield))
    for(i in 1:10){
        barley.lm <- lm(new.yield~variety+
                            year*site,weights=wt)
        wt <- wt.bisquare(barley.lm$res/median(abs(barley.lm$res)),c=6)
    }
    barley.effects <- dummy.coef(barley.lm)
    ans <- dotplot(sort(barley.effects$variety),
                   xlab = "Variety Effects (bushels/acre)", 
                   aspect=2/3)
    detach()
    ans
```
Now residuals
```{R}
    attach(barley)
    morris31 <- yield[(site=="Morris")&(year=="1931")]
    morris32 <- yield[(site=="Morris")&(year=="1932")]
    new.yield <- yield
    new.yield[(site=="Morris")&(year=="1931")] <- morris32
    new.yield[(site=="Morris")&(year=="1932")] <- morris31
    wt <- rep(1,length(yield))
    for(i in 1:10){
        barley.lm <- lm(new.yield~variety+year*site,weights=wt)
        wt <- wt.bisquare(barley.lm$res/median(abs(barley.lm$res)),c=6)
    }
    ans <- dotplot(site ~ residuals(barley.lm) | year * variety,
                   layout=c(4,5),
                   aspect=2/3,
                   panel = function(x, y) {
                       panel.dotplot(x, y)
                       panel.abline(v=0)
                   },
                   xlab = "Residual Barley Yield (bushels/acre)")
    detach()
    ans

    attach(barley)
    morris31 <- yield[(site=="Morris")&(year=="1931")]
    morris32 <- yield[(site=="Morris")&(year=="1932")]
    new.yield <- yield
    new.yield[(site=="Morris")&(year=="1931")] <- morris32
    new.yield[(site=="Morris")&(year=="1932")] <- morris31
    wt <- rep(1,length(yield))
    for(i in 1:10){
        barley.lm <- lm(new.yield~variety+year*site,weights=wt)
        wt <- wt.bisquare(barley.lm$res/median(abs(barley.lm$res)),c=6)
    }
    ans <- dotplot(variety ~ residuals(barley.lm) | year * site,
                   layout = c(2,6),
                   aspect=2/3,
                   panel = function(x, y) {
                       panel.dotplot(x, y)
                       panel.abline(v=0)
                   },
                   xlab = "Residual Barley Yield (bushels/acre)")
    detach()
    ans

```

