  ---
title: "Package `Reshape`"
author: "Song S. Qian"
date: "October 2, 2023"
output: pdf_document
---

# Front Matters


```{r, echo=FALSE, results='hide'}
source("FrontMatter.R")
packages(reshape)
```

# Package `reshape`

Package `reshape` provides a unified approach to aggregation, based on
an extended formula notation.  The core idea is to create a "melted"
version of a dataset using the `melt` function, which can then be cast
(using function `cast`) into an object with the desired orientation.
This means that we want to rearrange the form (not the content) of the
data to suit the need of analysis.

Before using the package, we need to familiarize the "anatomy" of a
dataset. We started by thinking of data in terms of vector, matrix, and
data frame.  In a data frame, observations are in rows and variables
are in columns.  To reshape a dataset, we can classify variables into
"identifier" and "measured" variables. 

An identifier (`id`) identifies the unit that the measurements take
place.  Id variables are usually discrete. A measured variable
represents what is measured on that unit, typically the response or
predictor variables. 

For example, in the `iris` dataset, we have fout measured variables
(`Sepal.Length, Sepal.Width, Petal.Length, Petal.Width`) and one id
variables (`Species`).  Later we reshaped the data to have three
columns, `Value, Variable`, and `Species`.  In this reshaped dataset,
the column `Value` is the measured variable and `Variables, Species`
are id variables.  In other words, we can generalize the data
classification to one measured variable and the rest are id variables.

## Melting data
The dataset with one column of measured variable values and id variables is the most basic data form which we can generate new forms. 

To produce the basic form of dataset is not a simple task.  The process of creating it is called melting.  The `reshape` package provides the function `melt` to carry out the task.  For example,  
```{r}
iris.melt <- melt(iris, id = "Species",
                  measured=c("Sepal.Width", "Sepal.Length", "Petal.Length","Petal.width"))
head(iris.melt)
```
The only assumption when using `melt` is that all measured variables
are in the same type (e.g., numeric, factor, date).

The dataset `iris.melt` is now a molten data, ready to be cast into
different forms.  In some cases, your datasets are already in the
molten form.  All you need to do is to make sure that the value column
is named `value`. 

## Casting molten data
We now have a molten dataset and we can cast it into different forms. All we need is to tell the function `cast` what form we need using a
formula. 

The casting formula describes the shape of the output format.  The general form of a formula includes two groups of variables separated by a tilde.  Variables on the left-hand side of the tilde are column variables and those on the right-hand side are row variables.  For example,  

```{r}
cast(iris.melt, Species ~ variable, length)
cast(iris.melt, Species ~ variable, mean)
```
Summarizes the sample size and mean.

The function `dcast` casts a molten data into data frame format.  The minimum input are (1) a molten data and (2) a formula.  The formula can also take $\cdot$ (a single dot) and $\cdots$ (three dots) to represent no variable and all other variables (not already included in the formula).

```{r}
cast(iris.melt, Species ~ variable)
cast(iris.melt, ... ~ variable)
cast(iris.melt, Species+variable ~ .)
```

What if we want to return to the original data frame?  In this case, the molten data lacks one piece of information:
```{r}
iris$rep <- rep(1:50, 3)
iris.melt2 <- melt(iris, id=c("Species", "rep"),
                   measured=c("Sepal.Width", "Sepal.Length", "Petal.Length","Petal.width"))
cast(iris.melt2, Species+rep ~ variable)
```

## The French Fries Example

```{r}
ffm <- melt(french_fries, id=1:4, na.rm=T)

### Counting non-missing values:
cast(ffm, .~., length)

### counting by treatment
cast(ffm, treatment ~ . , length)
cast(ffm, . ~ treatment, length)

### rep by treatment
cast(ffm, rep ~ treatment)

### treatment by rep:
cast(ffm, treatment ~ rep)

cast(ffm, treatment + rep ~ .)
cast(ffm, rep + treatment ~ .)
cast(ffm, . ~ rep+treatment)
```

Summary:
```{r}
cast(ffm, time+rep+subject+treatment + variable ~ .)
cast(ffm, rep+treatment~variable)
cast(ffm, rep+treatment~variable, mean)
cast(ffm, rep+treatment~variable, sd)
cast(ffm, treatment ~ ., c(mean, sd))
cast(ffm, treatment ~ ., summary)
cast(ffm, treatment ~ variable + result_variable, c(mean, sd))
cast(ffm, treatment + variable ~ result_variable, c(mean, sd))
```

Conditional operator

```{r}
cast(ffm, treatment~rep|variable, mean)
```
# Package `reshape2`

`reshape2` is a reboot of the `reshape` package, a new package for reshaping data that is much more focused and faster.

This version improves speed at the cost of functionality.

What's new in `reshape2`:

-- considerably faster and more memory efficient thanks to a much better underlying algorithm that uses the power and speed of
subsetting to the fullest extent, in most cases only making a single copy of the data. `cast`is replaced by two functions depending on the output type: `dcast` produces data frames, and `acast` produces matrices/arrays. 

-- multidimensional margins are now possible: `grand_row` and `grand_col` have been dropped: now the name of the margin refers to the variable that has its value set to (all).

-- some features have been removed such as the `cast` operator, and the ability to return multiple values from an aggregation function.

-- a new cast syntax which allows you to reshape based on functions of variables (based on the same underlying syntax as plyr): better development practices like namespaces and tests.
