---
title: "Reshaping and Merging"
author: "Song S. Qian"
date: "October 4, 2023"
output: pdf_document
---

# Front Matters

```{r, echo=FALSE, results='hide'}
source("../../R/FrontMatter.R")
packages(reshape)
```

# Reshaping Data
THe advantage of using R for statistical analysis is that R comes with
functions to carry out statistical analysis.  But each function
requires a specific input format.  As a result, reshaping data to suit
for the need of a function is the most commonly performed task. In
most cases, R functions require input using data frame.  We will focus
on working with data frames.

## Modifying data frame variables

Data frames are lists.  New variables can be creates by simply
assigning their value to a column that doesn't already exist in the
data frame.  Let's use the loblolly pine growth data as an example.
This dataset has three columns recording the growth of loblolly pine
trees over time.
```{r}
head(Loblolly)
```
Variable transformation is probably the most common modification to a dataset.  We add the transformed variable into an additional column:
```{r}
Loblolly$logheight <- log(Loblolly$height)
```
(When we use data from R or R packages, all changes we made will be
kept in the memory of the current R session.  Data from the system
won't be changed.) 

When the data frame name is long, we often use the function
`transform` to reduce the amount of typing.   
```{r}
Loblolly <- transform(Loblolly, logheight=log(height))
```

We access variables in a data frame through the data frame name plus
the `$` operator and variable names.  The function `with` helps to
reduce the repeated typing of the data frame name.  When using `with`,
R puts the data frame temperarily on the top of the search list: 
```{r}
logheight<-with(Loblolly, log(height))
```
When this line is submitted, R looks into the data frame `Loblolly` first to find `height`.  If a wrong name was typed, R would not be able to find it in the data frame, R will continue to search outside the data frame.
```{r}
height2 <- runif(20, 3, 85)
logheight2 <- with(Loblolly, log(height2))
```
We can remove a column of a data frame by setting its values to be `NULL`:
```{r}
head(Loblolly)
Loblolly$logheight <- NULL
head(Loblolly)
```
When a similar operation is needed for several columns and we want to
overwrite the original versions, we can put the multiple columns we
want to replace on the left hand side of the assignment operator.  For
example, if we want to convert the unit of iris measurements from
centimeters to inches, we need to divide each of the 4 measurement
variables by 2.54: 
```{r}
iris[,-5] <- sapply(iris[,-5], function(x) x/2.54)
```

Recoding a variable is another form of data transformation.  For
example, when the actual value of sepal length is not important, but
whether the length is above or below 6 cm is important, we would like
to have a column to indicate whether sepal length is above 6 or
not. This is a simple logic comparison: 
```{r}
largesepal <- iris$Sepal.Length > 6
```
In some cases, the function `ifelse` is very handy. The function takes
a logical vector as its first argument and two other arguments: the
first provides a value for the case where elements of the input
logical vector are true, and the second for the case where they are
false.  For example, suppose that we have experiments carried out in
10 sites and we realized that sites 2 and 4 can be grouped in one
group and the rest in another group.  The grouping can be created as
follows: 
```
newgroup <- ifelse(sites %in% c(2,4), 1, 2)
```
To make thing more complicated, suppose that sites 2 and 4 are in
group 1, sites 5 and 8 are in group 2 and the rest are in group 3: 
```
newgroup <- ifelse(sites %in% c(2,4), 1, 
                   ifelse (sites %in% c(5, 8), 2, 3))
```

## Reshaping a data frame
Not all data frame have all the necessary identifiers.  For example,
the `iris` data frame lacks the flower indentifier.  Let's work with
the iris data for one species: 
```{r}
iris.setosa <- iris[iris$Species=="setosa",-5]
head(iris.setosa)
```
Many R functions for statistical analysis requires that the data frame be reshaped to have a column of measured value and a column of variable.  
```{r}
setosaM <- melt(iris.setosa,
                measured=c("Sepal.Width", "Sepal.Length", "Petal.Length","Petal.width")) 
```
In this case, a much simpler approach would be to use the function `stack`:
```{r}
setosaS <- stack(iris.setosa)
```

The function `unstack` returns the data to its original form:
```{r}
setosaUS <- unstack(setosaS, values~ind)
```

## Function `reshape`
Not to be confused with the package `reshape`, the function `reshape`
is written for data with repeated measures to convert a data frame
between long and wide forms.  A long form data frame puts multiple
measurements of a subject (replicates) in separate rows, while the
wide form puts them in the same row.  Here is an example of long form: 
```{r}
obs <- data.frame(subj = rep(1:4, each = 3), 
                  time=rep(1:3, 4),
                  x=rnorm(12), y=rnorm(12))
head(obs)
```
A wide format has 4 rows, one for each subject:
```{r}
obs.wide <- reshape(obs, idvar="subj", v.names=c("x","y"), 
                         timevar="time", direction="wide")
obs.wide
str(obs.wide)

reshape(obs.wide, idvar="subj", direction="long")
```

## Combining Data Frames

The two frequently used functions are `rbind` (binding data frames by
row) and `cbind` (binding by column). 
```{r}
x <- data.frame(a=c("A","B","C"), b=c(1,2,3))
y <- data.frame(a=c("D","E","F","G"), b=c(3,4,5,6))

z <- rbind(x, y)
```
When using `rbind`, the two data frames must have the same names.  When using `cbind`, the two data frames must have the same number of rows (or one is a multipler of the other):
```{r}
cbind(y, z=c(1,2))
cbind(y, z=rep(1:2, 2))
```
But not:
```
cbind(x, z=c(1,2))
```

Merging two data frames based on a common factor is often necessary.
The function `merge` provides a flexible means for merging. 
```{r}
x<-data.frame(a=c(1,2,4,5,6), x=c(9,12,14,21,8))
y<-data.frame(a=c(1,3,4,6), y=c(8, 14, 19, 2))
merge(x, y)
```
The result may not be what you expect.  

By default, `merge` will find the common name in `x` and `y` to use as
the variable for merging (argument `by`).  If the merging variable has
different names in `x` and `y`, we can use argument `by.x` and `by.y`.
The argument `all` (default `all=FALSE`) is a logical value indicating
whether all rows in both `x` and `y` should be included: 
```{r}
merge(x, y, all=T)
```
If only all rows of `x` (`y`) are needed, we specify `all.x=T` (`all.y=T`).
```
merge(x, y, all.x=T)
merge(x, y, all.y=T)
```

One more example of `merge`:
```{r}
authors <- data.frame(
    surname = I(c("Tukey", "Venables", "Tierney", "Ripley", "McNeil")),
    nationality = c("US", "Australia", "US", "UK", "Australia"),
    deceased = c("yes", rep("no", 4)))
books <- data.frame(
    name = I(c("Tukey", "Venables", "Tierney",
             "Ripley", "Ripley", "McNeil", "R Core")),
    title = c("Exploratory Data Analysis",
              "Modern Applied Statistics ...",
              "LISP-STAT",
              "Spatial Statistics", "Stochastic Simulation",
              "Interactive Data Analysis",
              "An Introduction to R"),
    other.author = c(NA, "Ripley", NA, NA, NA, NA,
                     "Venables & Smith"))

m1 <- merge(authors, books, by.x="surname", by.y="name")
m2 <- merge(books, authors, by.x = "name", by.y = "surname")
m3 <- merge(books, authors, by.x = "name", by.y = "surname", all=T)
```

## Merging using index
The function `merge` is convenient.  In many cases, especially when we want to mearge data by a set of unique index (e.g., site names), we can simply use subscripting.  For example:
```{r}
x <- data.frame(site=c("S3","S2","S1"), Ag=c(20, 31, 19))
y <- data.frame(site=c("S1","S1","S1","S2","S2","S3","S3","S1","S2","S3"),
                y=rnorm(10))
```
Because the site names are the same, we can first sort `x` by site:
```{r}
x <- x[order(x$site),]
x
```
then:
```{r}
oo <- as.numeric(ordered(y$site))
oo
y$Ag <- x$Ag[oo]
y
```

# A quick review

## Data in R

Modes and classes, data storage in R, missing values

## Reading and writing data
Data frame: `read.table` and `read.csv`

Fixed width input files: `read.fwf`

Extracting data from R objects

Generating data -- sequences and random numbers

## Dates

`as.Date` function and the POSIX class

## Subscripting

Numeric, character, and logical subscripts, matrix, list, and data
frame (double versus single brackets) 

## Data aggregation

`table`, `apply`, `tapply`, `aggregate`, `by`, the `reshape` package
(`melt` and `cast`) 

## Data reshaping

Adding and subtracting columns from a data frame, recoding variables,
combining data frames (`cbind`, `rbind`, `merge`) 
