---
title: "EEES4160/5160 Homework 9"
output:
  pdf_document: default
  html_document:
    df_print: paged
date: "Due 11/17/2023"
Name: Lucas Finke
---

```{r setup, include=FALSE}
setwd("C:/Users/Lucas/Downloads/Kuhnert+Venables-R_Course_Notes/Session1/Session1/")
source("FrontMatter.R")
packages(tidyverse)
packages(lattice)
```

All homework assignments are to be completed using R in RStudio. The submitted version should be created using R Markdown. You should execute each line of code to check if they work properly before knitting into PDF (HTML, .docx). But do not include the print out the original data and the resulting subsets. There maybe multiple ways to complete the tasks.

This assignment is designed for practicing conditional plots using the `lattice` package function `xyplot`.  We will use the River Cam data.  The objective of these conditional plots is to visualize the data and understand how various factors affect the downstream DO concentration (variable `ddo`).

```{R Data Set, include=FALSE}
cam <- read.table("C:/Users/Lucas/Downloads/Kuhnert+Venables-R_Course_Notes/Session1/Session1/cam.dat",header=T)
head(cam)
```
Required package: `lattice` `r require(lattice)`

1. Using the figures of CP ratio against retinal area of cats in last week's class as examples to analyze the relationship between car's fuel efficiency and weight using the data `mtcars`. You should plot `mpg` against `wt` with a linear fit to the data. Then, use the residual plots (with a loess line) to check for linearity. Finally, plot `1/mpg` against `wt` and check residuals of the linear fit.

```{R mtcars, include=FALSE}
## require cats and car info
attach(mtcars)
```
The following sets up the first part of the first question -- plotting wt against mpg.
```{r wt against mpg}
book.3.2 <- function()
	xyplot(wt ~ mpg, 
		data = mtcars,
		panel = function(x,y){
			panel.xyplot(x,y)
			panel.loess(x,y, family = "g")
		},
		sub = list("Figure 3.2",cex=.8),
		xlab = "MPG",
		ylab = "wt",
		aspect="xy")

book.3.2()


ggVD_3.2<-
  function(){
    ggplot(mtcars, aes(mpg, wt))+geom_point(color="blue")+
        labs(x="MPG", y="wt")+
        geom_smooth(formula=y~x, se=FALSE, method="loess")
  }

ggVD_3.2()


```
This chunk sets up and shows the second part of the first question -- plotting wt to 1/mpg.

```{r wt verus 1/mpg}
book.3.2 <- function()
	xyplot(wt ~ 1/mpg, 
		data = mtcars,
		panel = function(x,y){
			panel.xyplot(x,y)
			panel.loess(x,y, family = "g")
		},
		sub = list("Figure 3.2",cex=.8),
		xlab = "1/MPG",
		ylab = "wt",
		aspect="xy")

book.3.2()


ggVD_3.2<-
  function(){
    ggplot(mtcars, aes(1/mpg, wt))+geom_point(color="blue")+
        labs(x="1/MPG", y="wt")+
        geom_smooth(formula=y~x, se=FALSE, method="loess")
  }

ggVD_3.2()

```
1. The first plot we discussed in class is how the effect of sunlight hour varies as a function of temperature.  We saw scatter plots of `ddo` against `sun` conditional on temperature (`temp`).  Use the same code in the lecture notes, but change the number of intervals of the conditioning variable (`temp`). ## changed to 3.
The goal is to select the number of intervals that can best illustrate the interaction between `sun` and `temp`.  The visualization can also be more effective using different layout.  Use the option `layout` in `xyplot` to select the most effective presentation.  

This next chunk plots the sun vs ddo, using the data from the river camera.

```{r plotting ddo and sun}
xyplot(ddo ~ sun, data=cam) 
```

This next portion uses diffferent intervals for the xyplot of ddo and sun in respects to the temperature. 

```{r ddo intervals} 
tmp <- equal.count(cam$temp, 4, 0.25)
xyplot(ddo ~ sun | tmp, data = cam,
        xlab = "Sunlight hour", ylab = "Downstream DO (mg/L)",
        panel = function(x, y) {
            panel.grid(h=-1, v= 2)
            panel.xyplot(x, y)
            panel.loess(x,y, span=1)
        },
        aspect = 1)
```
There is always a sunlight effect, but the effect is obvious only when the temperature is high.

2. Repeat exercise 1 by using temperature as the conditioning variable (`xyplot(ddo~temp|snlgt)`). 

```{R intervals sunlight}
snlgt <- equal.count(cam$sun, 6, 0.25)
xyplot(ddo ~ temp | snlgt, data = cam,
        xlab = "Temperature (C)", ylab = "Downstream DO (mg/L)",
        panel = function(x, y) {
            panel.grid(h=-1, v= 2)
            panel.xyplot(x, y)
            panel.loess(x,y, span=1)
        }, layout=c(3,2),
        aspect = 1)
```

3. Redraw the final plots of exercises 1 and 2 by replacing the line `panel.loess(x,y,span=1)` with `panel.lmline(x,y)`. 

The difference between the two plots are that the lmline plots are more linearly regressed than the loess plots. the number of windows were changed as well. 

```{r intervals temp}
tmp <- equal.count(cam$temp, 3, 0.25)
xyplot(ddo ~ sun | tmp, data = cam,
        xlab = "Sunlight hour", ylab = "Downstream DO (mg/L)",
        panel = function(x, y) {
            panel.grid(h=-1, v= 2)
            panel.xyplot(x, y)
            panel.lmline(x,y, span=1)
        },
        aspect = 1)
```

This next chunk uses almost the same exact code as the previous one, but comparing ddo to temperature in respects to sunlight at a different interval. 

```{R sunlight intervals}
snlgt <- equal.count(cam$sun, 3, 0.25)
xyplot(ddo ~ temp | snlgt, data = cam,
        xlab = "Temperature (C)", ylab = "Downstream DO (mg/L)",
        panel = function(x, y) {
            panel.grid(h=-1, v= 2)
            panel.xyplot(x, y)
            panel.lmline(x,y, span=1)
        }, layout=c(3,2),
        aspect = 1)
```
4. The main goal of MB Beck (who collected the data) was to understand the DO and BOD dynamics.  Temperature and sunlight hour change this relationship because algal biomass in water will be represented in BOD.  In class, we used `xyplot` study the `ddo~dbod` relationship conditioned on temperature and sunlight.  Explore the number of intervals that can best illustrate the effect of these two conditioning variables.


```{r intervals sunltemp}
snlgt <- equal.count(cam$sun, 3, 0.25)
tmp <- equal.count(cam$temp, 3, 0.25)
xyplot(ddo ~ dbod | snlgt*tmp, data = cam,
        xlab = "Downstream BOD", ylab = "Downstream DO (mg/L)",
        panel = function(x, y) {
            panel.grid(h=-1, v= 2)
            panel.xyplot(x, y)
##            panel.lmline(x,y)
            panel.loess(x, y, span=1)
        },
        aspect = 1)
```


5. Repeat exercise 4 by replacing variable `dbod` with `ubod`.  Upstream BOD measures the input organic matter pollution.  Downstream BOD represents both the remaining organic matter from upstream and the algal biomass produced in the study river reach.

```{R ubod intervals}
xyplot(ddo ~ ubod | snlgt*tmp, data = cam,
        xlab = "Downstream BOD", ylab = "Downstream DO (mg/L)",
        panel = function(x, y) {
            panel.grid(h=-1, v= 2)
            panel.xyplot(x, y)
##            panel.lmline(x,y)
            panel.loess(x, y, span=1)
        },
        aspect = 1)
```

